---
title: "BIO 611 Final Project (Alzheimer's Mortality & Risk Factors in US)"
name: "Lauren Young"
Date: "12 May 2026"
output: html_notebook
---

###LOAD LIBRARIES & READ IN DATA
```{r message = "FALSE", warning = "FALSE"}
library(tidyverse)
library(dplyr)
library(patchwork)
library(car)
library(ggplot2)
library(reshape2)
library(scales)
library(ggfortify)
library(plotly)
library(ggrepel)
library(lmtest)
library(htmlwidgets)
library(htmltools)
library(corrplot)
library(broom)
```

###READ IN ALL DATA
```{r results = "hide"}
#NCHS disease mortality data
NCHS <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/NCHS_-_Leading_Causes_of_Death__United_States.csv")

#read in census ACS 2016 5-yr estimate median income data
income <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/Census Data/ACSST5Y2016.S1903_2026-04-15T140659/ACSST5Y2016.S1903-Data.csv")

#read in census ACS 2016 5-yr estimate education data
education <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/Census Data/ACSST5Y2016.S1501_2026-05-06T215714/ACSST5Y2016.S1501-Data.csv") 

#read in census ACS 5-yr estimate health insurance data
insurance <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/Census Data/ACSST5Y2016.S2701_2026-05-07T103058/ACSST5Y2016.S2701-Data.csv") 

#read in census ACS 5-yr estimate poverty data
poverty <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/Census Data/ACSST5Y2016.S1701_2026-05-08T074102/ACSST5Y2016.S1701-Data.csv")

#read in census ACS 5-yr estimtate age data
age <- read.csv("C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Data/Census Data/ACSST5Y2016.S0101_2026-04-15T140727/ACSST5Y2016.S0101-Data.csv") 
```

###TIDY/CLEAN NCHS DATA
```{r results = "hide", warning = "FALSE"}
#filter for alzheimers disease
NCHS_alzheimers <- NCHS %>% filter(Cause.Name == "Alzheimer's disease")
glimpse(NCHS_alzheimers)

#remove US total row 
NCHS_final <- NCHS_alzheimers %>%
  filter(State != "United States") %>%
  
  filter(Year == 2017) %>% #filter for 2017 year
  
  rename(death_rate = Age.adjusted.Death.Rate) %>% #rename death rate column
  
  select(-X113.Cause.Name) #drop second cause name column

#convert deaths and death rate to numeric data
NCHS_final <- NCHS_final %>%
  mutate(
    Deaths = as.numeric(gsub(",", "", as.character(Deaths))),
    death_rate = as.numeric(as.character(death_rate)))

summary(NCHS_final)
```

###Create state-fips lookup table
```{r}
state_lookup <- data.frame(
  state_fips = c(
    "01","02","04","05","06","08","09","10","11","12",
    "13","15","16","17","18","19","20","21","22","23",
    "24","25","26","27","28","29","30","31","32","33",
    "34","35","36","37","38","39","40","41","42","44",
    "45","46","47","48","49","50","51","53","54","55","56"
  ),
  State = c(
    "Alabama","Alaska","Arizona","Arkansas","California",
    "Colorado","Connecticut","Delaware","District of Columbia",
    "Florida","Georgia","Hawaii","Idaho","Illinois","Indiana",
    "Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland",
    "Massachusetts","Michigan","Minnesota","Mississippi",
    "Missouri","Montana","Nebraska","Nevada","New Hampshire",
    "New Jersey","New Mexico","New York","North Carolina",
    "North Dakota","Ohio","Oklahoma","Oregon","Pennsylvania",
    "Rhode Island","South Carolina","South Dakota","Tennessee",
    "Texas","Utah","Vermont","Virginia","Washington",
    "West Virginia","Wisconsin","Wyoming"))
```


###TIDY/CLEAN INCOME 
```{r results = "hide", warning = "FALSE"}
#convert income/household columns to numeric
income2 <- income %>%
  mutate(
    S1903_C02_001E = parse_number(S1903_C02_001E),
    S1903_C01_001E = parse_number(S1903_C01_001E))

#extract state fips codes
income3 <- income2 %>%
  mutate(state_fips = substr(GEO_ID, 10, 11))

#aggregate to state level
income4 <- income3 %>%
  group_by(state_fips) %>%
  summarise(
    median_income = mean(S1903_C02_001E, na.rm = TRUE),
    households = sum(S1903_C01_001E, na.rm = TRUE))

#remove first and last row
income5 <- income4 %>%
  
  filter(!is.na(state_fips),
         state_fips != "") %>%
  
  filter(!is.na(state_fips),
         state_fips != "72") %>%
  
  mutate(state_fips = as.character(state_fips))   #check that state fips is character

#add state names
income_final <- income5 %>%
  left_join(state_lookup, by = "state_fips")

```

###TIDY/CLEAN EDUCATION
```{r results = "hide", warning = "FALSE"}
#convert education column to numeric
ed2 <- education %>%
  mutate(S1501_C02_015E = parse_number(S1501_C02_015E)) #percent of pop with bachelors degree or higher

#extract state fips codes
ed3 <- ed2 %>%
  mutate(state_fips = substr(GEO_ID, 10, 11))

#aggregate to state level
ed4 <- ed3 %>%
  group_by(state_fips) %>%
  summarise(
    bachelor = mean(S1501_C02_015E, na.rm = T))

#remove first and last row
ed5 <- ed4 %>%
  
  filter(!is.na(state_fips),
         state_fips != "") %>%
  
  filter(!is.na(state_fips),
         state_fips != "72") %>%
  
  mutate(state_fips = as.character(state_fips))   #check that state fips is character

#add state names
ed_final <- ed5 %>%
  left_join(state_lookup, by = "state_fips")
```

###TIDY/CLEAN HEALTH INSURANCE
```{r results = "hide", warning = "FALSE"}
#convert health insurance column to numeric
ins2 <- insurance %>%
  mutate(S2701_C05_001E = parse_number(S2701_C05_001E)) #population uninsured (%)

#extract state fips codes
ins3 <- ins2 %>%
  mutate(state_fips = substr(GEO_ID, 10, 11))

#aggregate to state level
ins4 <- ins3 %>%
  group_by(state_fips) %>%
  summarise(
    uninsured = mean(S2701_C05_001E, na.rm = T))

#remove first and last row
ins5 <- ins4 %>%
  
  filter(!is.na(state_fips),
         state_fips != "") %>%
  
  filter(!is.na(state_fips),
         state_fips != "72") %>%
  
  mutate(state_fips = as.character(state_fips))   #check that state fips is character

#add state names
ins_final <- ins5 %>%
  left_join(state_lookup, by = "state_fips")
```

###TIDY/CLEAN POVERTY
```{r reaults = "hide", warning = "FALSE"}
#convert poverty column to numeric
pov2 <- poverty %>%
  mutate(S1701_C03_001E = parse_number(S1701_C03_001E)) #population in poverty (%) #total population

#extract state fips codes
pov3 <- pov2 %>%
  mutate(state_fips = substr(GEO_ID, 10, 11))

#aggregate to state level
pov4 <- pov3 %>%
  group_by(state_fips) %>%
  summarise(
    poverty = mean(S1701_C03_001E, na.rm = T))

#remove first and last row
pov5 <- pov4 %>%
  
  filter(!is.na(state_fips),
         state_fips != "") %>%
  
  filter(!is.na(state_fips),
         state_fips != "72") %>%
  
  mutate(state_fips = as.character(state_fips))   #check that state fips is character

#add state names
pov_final <- pov5 %>%
  left_join(state_lookup, by = "state_fips")
```

###TIDY/CLEAN AGE 
```{r}
#convert age column to numeric
age2 <- age %>%
  mutate(S0101_C01_028E = parse_number(S0101_C01_028E)) #population >65 (%) #total population

#extract state fips codes
age3 <- age2 %>%
  mutate(state_fips = substr(GEO_ID, 10, 11))

#aggregate to state level
age4 <- age3 %>%
  group_by(state_fips) %>%
  summarise(
    age = mean(S0101_C01_028E, na.rm = T))

#remove first and last row
age5 <- age4 %>%
  
  filter(!is.na(state_fips),
         state_fips != "") %>%
  
  filter(!is.na(state_fips),
         state_fips != "72") %>%
  
  mutate(state_fips = as.character(state_fips))   #check that state fips is character

#add state names
age_final <- age5 %>%
  left_join(state_lookup, by = "state_fips")
```

###CREATE FINAL DATASET (MERGE NCHS & SOCIAL-ECONOMIC FACTORS)
```{r}
data <- merge(NCHS_final, income_final, by = c("State"), all = T)

data2 <- merge(data, ed_final, by = c("State", "state_fips"), all = T)

data3 <- merge(data2, ins_final, by = c("State", "state_fips"), all = T)

data4 <- merge(data3, pov_final, by = c("State", "state_fips"), all = T)

data_final <- merge(data4, age_final, by = c("State", "state_fips"), all = T)

summary(data_final)
```

###DESCRIPTIVE STATISTICS
```{r}
hist(data_final$death_rate)
hist(data_final$median_income)
hist(data_final$bachelor)
hist(data_final$uninsured)
hist(data_final$poverty)
hist(data_final$age)

boxplot(data_final$death_rate)

shapiro.test(data_final$death_rate)

```

###MULTIPLE LINEAR REGRESSION
```{r}
#standardize variables
data_scaled <- data_final %>%
  mutate(across(c(death_rate, 
                  median_income, 
                  bachelor, 
                  uninsured, 
                  poverty,
                  age), scale))

#fit model
model <- lm(death_rate ~ 
              median_income + 
              bachelor + 
              uninsured + 
              poverty +
              age,
            data = data_scaled)

summary(model)

#check model assumptions
cor(data_final[, c("death_rate", "median_income", "bachelor", "uninsured", "poverty", "age")])

vif(model)

shapiro.test(residuals(model))

par(mfrow = c(2,2))
plot(model)

```

#####VISUALIZATIONS

###SCATTERPLOTS

#Median Income
```{r results = "hide"}
#regression for income and death rate
model_income <- lm(death_rate ~ median_income, data = data_final)

#use predict function
predict(model_income)

#make new income values
new.income <- expand.grid(median_income = seq(from = 35000, to = 85000, length.out = 100))
new.income

#generate fits and CIs at new income values
new.y.income <- predict(model_income, newdata = new.income, interval = "confidence")
new.y.income

#create predicted values
PredictValues.income <- data.frame(new.income, new.y.income)
PredictValues.income <- rename(PredictValues.income, death_rate = fit)
PredictValues.income
```

```{r}
#plot predicted values
Income_figure <- ggplot(data_final, aes(x = median_income, y = death_rate)) +
  geom_point(size = 3, colour = "black") +
  geom_smooth(data = PredictValues.income, aes(ymin = lwr, ymax = upr), stat = "identity", colour = "#66CD00") +
  xlab("Median income") +
  ylab("Age-adjusted death rates (per 100,000 population)") +
  scale_x_continuous(labels = dollar_format(prefix = "$", big.mark = ",")) +
  theme_classic(base_size = 24)

Income_figure
```

#Education
```{r results = "hide"}
#regression for education and death rate
model_ed <- lm(death_rate ~ bachelor, data = data_final)

#use predict function
predict(model)

#make new income values
new.ed <- expand.grid(bachelor = seq(from = 10, to = 60, length.out = 5))
new.ed

#generate fits and CIs at new income values
new.y.ed <- predict(model_ed, newdata = new.ed, interval = "confidence")
new.y.ed

#create predicted values
PredictValues.ed <- data.frame(new.ed, new.y.ed)
PredictValues.ed <- rename(PredictValues.ed, death_rate = fit)
PredictValues.ed
```

```{r}
#plot predicted values
Education_figure <- ggplot(data_final, aes(x = bachelor, y = death_rate)) +
  geom_point(size = 3, colour = "black") +
  geom_smooth(data = PredictValues.ed, aes(ymin = lwr, ymax = upr), stat = "identity", colour = "#4292c6") +
  xlab("Bachelor's degree or higher (%)") +
  ylab("Age-adjusted death rates (per 100,000 population)") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme_classic(base_size = 24)

Education_figure

```

#Insurance
```{r results = "hide"}
#regression for insurance and death rate
model_ins <- lm(death_rate ~ uninsured, data = data_final)

#use predict function
predict(model_ins)

#make new income values
new.ins <- expand.grid(uninsured = seq(from = 0, to = 25, length.out = 2))
new.ins

#generate fits and CIs at new income values
new.y.ins <- predict(model_ins, newdata = new.ins, interval = "confidence")
new.y.ins

#create predicted values
PredictValues.ins <- data.frame(new.ins, new.y.ins)
PredictValues.ins <- rename(PredictValues.ins, death_rate = fit)
PredictValues.ins
```

```{r}
#plot predicted values
Insurance_figure <- ggplot(data_final, aes(x = uninsured, y = death_rate)) +
  geom_point(size = 3, colour = "black") +
  geom_smooth(data = PredictValues.ins, aes(ymin = lwr, ymax = upr), stat = "identity", colour = "#FF7F00") +
  xlab("Population uninsured (%)") +
  ylab("Age-adjusted death rates (per 100,000 population)") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme_classic(base_size = 24)

Insurance_figure
```

#Poverty
```{r results = "hide"}
#regression for poverty and death rate
model_pov <- lm(death_rate ~ poverty, data = data_final)

#use predict function
predict(model_pov)

#make new income values
new.pov <- expand.grid(poverty = seq(from = 6, to = 22, length.out = 2))
new.pov

#generate fits and CIs at new income values
new.y.pov <- predict(model_pov, newdata = new.pov, interval = "confidence")
new.y.pov

#create predicted values
PredictValues.pov <- data.frame(new.pov, new.y.pov)
PredictValues.pov <- rename(PredictValues.pov, death_rate = fit)
PredictValues.pov
```

```{r}
#plot predicted values
Poverty_figure <- ggplot(data_final, aes(x = poverty, y = death_rate)) +
  geom_point(size = 3, colour = "black") +
  geom_smooth(data = PredictValues.pov, aes(ymin = lwr, ymax = upr), stat = "identity", colour = "#8B5742") + 
  xlab("Population in poverty (%)") +
  ylab("Age-adjusted death rates (per 100,000 population)") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme_classic(base_size = 24)

Poverty_figure
```

#Age
```{r results = "hide"}
#regression for age and death rate
model_age <- lm(death_rate ~ age, data = data_final)

#use predict function
predict(model_age)

#make new income values
new.age <- expand.grid(age = seq(from = 8, to = 20, length.out = 2))
new.age

#generate fits and CIs at new income values
new.y.age <- predict(model_age, newdata = new.age, interval = "confidence")
new.y.age

#create predicted values
PredictValues.age <- data.frame(new.age, new.y.age)
PredictValues.age <- rename(PredictValues.age, death_rate = fit)
PredictValues.age
```

```{r}
#plot predicted values
Age_figure <- ggplot(data_final, aes(x = age, y = death_rate)) +
  geom_point(size = 3, colour = "black") +
  geom_smooth(data = PredictValues.age, aes(ymin = lwr, ymax = upr), stat = "identity", colour = "#EEEE00") + 
  xlab("Population above 65 years old (%)") +
  ylab("Age-adjusted death rates (per 100,000 population)") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme_classic(base_size = 24)

Age_figure
```

#Final Scatterplot
```{r}
Figure1 <- (Income_figure + Education_figure + Insurance_figure +
            Poverty_figure + Age_figure) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = "A")


Figure1 <- (Income_figure + Education_figure + Insurance_figure +
            Poverty_figure + Age_figure) +
  plot_layout(nrow = 2) +
  plot_annotation(tag_levels = "A") &
  theme(
    text = element_text(size = 30), 
    axis.title = element_text(size = 30),    
    axis.text = element_text(size = 26),     
    plot.tag = element_text(size = 30))

Figure1

ggsave("Figure1.png", width = 30, height = 25)
```


###MAPS

#Death Rate
```{r}
#remove DC
Mortality_map <- data_final %>% filter(State != "District of Columbia")

#lowercase state names
Mortality_map$region <- tolower(Mortality_map$State)

#get US map data
us_map <- map_data("state")

#join with US map data
Mortality_map_joined <- us_map %>%
  left_join(Mortality_map, by = "region")

#create tooltip text
Mortality_map_joined$tooltip <- paste0(
  Mortality_map_joined$State,
  "<br>Death rate: ",
  comma(Mortality_map_joined$death_rate))

#create map
Mortality_map_final <- ggplot(
  Mortality_map_joined,
  aes(
    long,
    lat,
    group = group,
    fill = death_rate,
    text = tooltip)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(
     low = "#F2F0F7",
    high = "#54278F",  
    na.value = "grey90") +
  labs(fill = "Death rate") +
  theme_void(base_size = 24)

Mortality_map_interact <- ggplotly(Mortality_map_final, tooltip = "text")

Mortality_map_interact

# save interactive plot as a .html file
htmlwidgets::saveWidget(Mortality_map_interact,"C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Mortality_map_interact.html")

```

#Median Income
```{r}
#remove DC
Income_map <- data_final %>% filter(State != "District of Columbia")

#lowercase state names
Income_map$region <- tolower(Income_map$State)

#get US map data
us_map <- map_data("state")

#join with US map data
Income_map_joined <- us_map %>%
  left_join(Income_map, by = "region")

#create tooltip text
Income_map_joined$tooltip <- paste0(
  Income_map_joined$State,
  "<br>Median income: $",
  comma(Income_map_joined$median_income))

#create map
Income_map_final <- ggplot(
  Income_map_joined,
  aes(
    long,
    lat,
    group = group,
    fill = median_income,
    text = tooltip)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(
    low = "#e5f5e0",  
    high = "#006400",  
    na.value = "grey90",
    labels = dollar_format(prefix = "$", big.mark = ",", accuracy = 1)) +
  labs(fill = "Median income") +
  theme_void(base_size = 24)

Income_map_interact <- ggplotly(Income_map_final, tooltip = "text")
```

#Age
```{r}
#remove DC
Age_map <- data_final %>% filter(State != "District of Columbia")

#lowercase state names
Age_map$region <- tolower(Age_map$State)

#get US map data
us_map <- map_data("state")

#join with US map data
Age_map_joined <- us_map %>%
  left_join(Age_map, by = "region")

#create tooltip text
Age_map_joined$tooltip <- paste0(
  Age_map_joined$State,
  "<br>Age 65 + (%): %",
  comma(Age_map_joined$age))


#create map
Age_map_final <- ggplot(
  Age_map_joined,
  aes(
    long,
    lat,
    group = group,
    fill = age,
    text = tooltip)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
scale_fill_gradient(
  low = "#FFFFE5",
  high = "#CDAD00",
  labels = function(x) paste0(x, "%")) +
  labs(fill = "Age 65+ (%)") +
  theme_void(base_size = 24)

Age_map_interact <- ggplotly(Age_map_final, tooltip = "text")

Age_map_interact

# save interactive plot as a .html file
htmlwidgets::saveWidget(Age_map_interact,"C:/Users/laure/OneDrive - James Madison University/BIO 611/Project/Age_map_interact.html")

```

###CORRELATION HEATMAP
```{r}
vars <- data_scaled %>%
  select(death_rate, median_income, bachelor, uninsured, poverty, age)

colnames(vars) <- c(
  "Death rate",
  "Median income",
  "Bachelor's degree\nor higher (%)",
  "Uninsured (%)",
  "Poverty (%)",
  "Age 65+ (%)")

cor_matrix <- cor(vars, use = "complete.obs")

png("correlation_plot.png",
    width = 2000,
    height = 1600,
    res = 300)

correlation_plot <- corrplot::corrplot(cor_matrix,
         method = "color",
         type = "upper",
         order = "hclust",
         col = colorRampPalette(c("#B2182B", "white", "#3A5FCD"))(200),
         addCoef.col = "black",
         tl.col = "black",
         tl.cex = 1.1,
         tl.srt = 45,
         tl.offset = 0.8,
         number.cex = 1,
         cl.cex = 1,
         diag = FALSE)

title(main = "A", adj = 0, cex.main = 1.5, font.main = 2)

while (!is.null(dev.list())) {
 dev.off()}

```

###COEFFICIENT PLOT
```{r}
tidy_model <- tidy(model)

tidy_model$term <- dplyr::recode(tidy_model$term,
  median_income = "Median income",
  bachelor = "Bachelor's degree or higher (%)",
  uninsured = "Uninsured (%)",
  poverty = "Poverty (%)",
  age = "Age 65+ (%)")

#extract R-sqaured values
model_r2 <- summary(model)$r.squared

#create plot
coefficient_plot <- ggplot(
  tidy_model,
  aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "#BDBDBD") +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = estimate - 1.96 * std.error,
      ymax = estimate + 1.96 * std.error),
    width = 0.2) +
  coord_flip() +
  theme_classic(base_size = 30) +
  labs(
    x = "",
    y = "Effect size (95% CI)",
    tag = "B") +
  theme(
    plot.tag = element_text(size = 32, face = "bold"),
    plot.tag.position = c(0.07, 0.94)) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = paste0("R² = ", round(model_r2, 3)),
    hjust = 1.1,
    vjust = 1.5,
    size = 8)

ggsave("coefficient_plot.png", width = 20, height = 10)

```


